Ages of Planets and Stars Observed by Kepler

### 2024-04-20 paper/ms.tex
ms: text improvements

% number of stars monitored by Kepler (quoting Santos19/21)
\newcommand{\nkeplerstars}{$\approx$160{,}000}
% fraction of stars with rotation periods, with teff/logg in Berger+20
\newcommand{\fracstarswithprotwithbtwenty}{{$\sim$94\%}}
% complement of above
\newcommand{\fracstarswithprotwithoutbtwenty}{{$\sim$6\%}}

%%%%%%%%%%%
% PLANETS %
%%%%%%%%%%%
% number of KOIs in Q1-Q17 DR25 table
%\newcommand{\nkois}{{$\sim$8{,}000}} %FIXME precise
% number of KOIs in Q1-Q17 DR25 table that are not known false positives
%\newcommand{\nkoisnofp}{{$\sim$4{,}000}} %FIXME precise
% number of KOIs in Q1-Q17 DR25 table that have rotation periods
\newcommand{\nkoiswithprot}{{$\sim$2{,}000}}
% number of KOIs in Q1-Q17 DR25 table that are not known false positives that have rotation periods
\newcommand{\nkoisnofpwithprot}{{$\sim$1{,}000}}
% fraction of KOIs that are not FPs with B+20 parameters
\newcommand{\frackoisnofpwithprotwithbtwenty}{{$\sim$92\%}}

@@

Discovering a young planet requires solving two problems: find the
planet, and measure the star's age.  Each problem admits a range of
solutions.  Here, we will focus on planets found using transits,
and stellar ages inferred using rotation and lithium.  Rotation
will provide the most age information for the largest number of
stars.  The lithium-based technique will provide useful information
for stars younger than a few hundred million years.

To begin, let us imagine that the age distribution of stars within a
few hundred parsecs is uniformly distributed between 0 and 10\,Gyr
\citep[e.g.][]{Nordstrom_2004}.  
The utility of this approximation is that it provides a rough estimate
for the rarity of young stars:
$\approx$1\% of stars in the nearby Galaxy have ages below 100\,Myr,
and $\approx$10\% are younger than 1\,Gyr.
Observations of exoplanets forming in disks (CITE), and of exoplanets
evolving shortly after disk dispersal (CITE CITE), are therefore
severely limited in their achievable sample sizes by the
tyranny of the Galactic star formation rate.

@@

This work is focused on Kepler stars for which ages can be inferred
using either stellar rotation, lithium, or both.  These ``age-dateable
stars'' are a minority of the \nkeplerstars\ Kepler targets.  Rotation
periods provide the majority of the information, since they have been
reported for roughly one in three Kepler targets
\citep[e.g.][]{McQuillan_2014,Santos_2021}.  High-resolution spectra,
and therefore measurements of the \ion{Li}{1} 6708\,\AA\ doublet, in
general have only been acquired for the Kepler objects of interest
(KOIs), which comprise only a few percent of the overall stellar
sample.  In the following, we will describe the set of stars that we
adopt with measured rotation periods (Section~\ref{subsec:rotsel}),
the set of objects we adopt as planets
(Section~\ref{subsec:planetsel}), and the set of planets with
high-resolution spectra suitable for lithium analysis
(Section~\ref{subsec:lithiumsel}).

@@

We considered planets in the NASA Exoplanet Archive's cumulative KOI
table, which included the best knowledge available on any given planet
candidate as of 6 June 2023, while also incorporating human-based
vetting \footnote{See
\url{https://exoplanetarchive.ipac.caltech.edu/docs/PurposeOfKOITable.html}}.
These planets represent a superset of those in the fully automated
Q1-Q17 DR25 KOI Table \citep{Thompson_2018}, which could be adopted in
future work for planet occurrence rate calculations.  This version of
the cumulative KOI table included \nkoisnofp\ that are either
``confirmed'' or ``candidate'' planets, after excluding known false
positives.

@@

\subsection{High Resolution Spectra}
A final piece of our analysis involves an age assessment based on the
equivalent width of the \ion{Li}{1} 6708\,\AA\ doublet.  For this
component of the work, we rely on spectra acquired with the High
Resolution Echelle Spectrometer (HIRES; \citealt{vogt_hires_1994}) on
the Keck I 10m telescope.  These spectra were collected between 2009
and 2024, primarily through projects such as the California Kepler
Survey
\citep{2017AJ....154..107P,2017AJ....154..108J,2017AJ....154..109F}.
We supplemented the existing archive with $\approx$10\,hours of new
observations for 22 stars between Fall 2022 and Spring 2024.
These targets were selected to ensure that confirmed planets with
evidence from rotation for ages below 1\,Gyr received spectra, since
this is the age range where lithium was most likely to yield
useful age constraints.

@@

As our default source for stellar temperatures and surface gravities,
we adopt the values from the Gaia-Kepler Stellar Properties Catalog
(GKSP; \citealt{Berger_2020a_catalog}).  The GKSP parameters are
available for stars with ``AAA'' 2MASS photometry, measured parallaxes
in Gaia DR2,  and $g$-band photometry from either the KIC or the
Kepler-INT survey.  The parameters themselves were derived using
\texttt{isoclassify} \citep{2017ApJ...844..102H} to interpolate over
the MIST isochrone grids
\citep{2016ApJ...823..102C,2016ApJS..222....8D}, given the SDSS $g$
and 2MASS $K_{\rm s}$ photometry, the Gaia DR2 parallaxes, and
$V$-band extinction from the \citet{2018MNRAS.478..651G} reddening
map.  The resulting stellar parameters are available for
\fracstarswithprotwithbtwenty\ of the \nuniqstarsantosrot\ Kepler
stars with rotation periods.
\citet{David_2021} compared the stellar parameters ($T_{\rm eff}$,
$R_\star$, [Fe/H]) derived from the \citet{Berger_2020a_catalog}
isochronal approach with the spectroscopic parameters from
\citet{Fulton_2018}.  While this approach is one way of validating the
\citet{Berger_2020a_catalog} output, the temperature scales did show a
few-percent difference, with \citetalias{Fulton_2018} quoting higher
temperatures than \citetalias{Berger_2020a_catalog} for mid K dwarfs,
and lower temperatures for early F dwarfs.

%For the remaining \fracstarswithprotwithoutbtwenty\ of stars that lack
%temperatures and surface gravities, we adopt the values 
%from the Gaia DR3 General Stellar Parametrizer from Photometry (GSP-Phot; CITE).
%GSP-Phot uses the low-resolution Gaia BP and RP spectra, mean $G$ magnitudes,
%and parallaxes, in concert with the PARSEC isochrones (CITE Tang2014, Chen2015,
%Pastorelli2020) to infer the stellar effective temperatures, surface gravities,
%metallicities, and line-of-sight extinctions.
%In the planet sample,
%\frackoisnofpwithprotwithbtwenty\ of the
%the non-false-positive KOIs with rotation periods have parameters from
%\citet{Berger_2020a_catalog}.

For the remaining \fracstarswithprotwithoutbtwenty\ of stars that lack
temperatures and surface gravities from
\citetalias{Berger_2020a_catalog}, we adopt the values reported by
\citet{Santos_2019} and \citet{Santos_2021}, which for these cases
primarily derive from the \citet{Mathur_2017} DR25 Kepler Stellar
Properties Catalog, and are mostly derived from photometry.  In the
planet sample, \frackoisnofpwithprotwithbtwenty\ of the
non-false-positive KOIs with rotation periods have parameters from
\citet{Berger_2020a_catalog}, and the remainder similarly are drawn
from DR25.

%TODO: the relevant check here is Berger20 teff vs CKS spec teff...
%  As a sanity check on our effective temperature scale, we then used the 
%  the CITE Green2019 reddening map, combined with the Gaia DR2 $G_{\rm
%  BP}-G_{\rm RP}$ color, to calculate effective temperatures using the
%  relation suggested by \citet{Curtis_2020}.
%  This latter effective temperature scale was demonstrated to have a 
%  scatter of $\approx$50\,K for stars near the main-sequence.
%  We find {\bf X, Y, Z.}
%  We ultimately favor the \citet{Berger_2020a_catalog} scale
%  because it uses 2MASS $K_{\rm s}$ photometry, and is tied to an actual
%  isochronal fit.

%TODO: plot uncertainties in teff/logg for both parameter sets.

% TODO: add that plot showing differences btwn the two teff scales?

@@

To evaluate the accuracy and precision of our equivalent measurement
technique, after applying an initial iteration on the
\nplhoststarwgyroagewithgrazingandhighruwe\ aforementioned stars,
we compared our lithium equivalent widths
with those reported by \citet{Berger_2018}.
For the \nbergeroverlap\ stars in both samples, we found {\bf X, Y,
Z.}

>>>>

% number of stars monitored by Kepler (quoting Santos19/21)
\newcommand{\nkeplerstars}{$\approx$160{,}000}
% fraction of stars with rotation periods, with teff/logg in Berger+20
\newcommand{\fracstarswithprotwithbtwenty}{{$\approx$94\%}}
% complement of above
\newcommand{\fracstarswithprotwithoutbtwenty}{{$\approx$6\%}}

%%%%%%%%%%%
% PLANETS %
%%%%%%%%%%%
% number of KOIs in Q1-Q17 DR25 table
%\newcommand{\nkois}{{$\sim$8{,}000}} %FIXME precise
% number of KOIs in Q1-Q17 DR25 table that are not known false positives
%\newcommand{\nkoisnofp}{{$\sim$4{,}000}} %FIXME precise
% number of KOIs in Q1-Q17 DR25 table that have rotation periods
\newcommand{\nkoiswithprot}{{$\sim$2{,}000}}
% number of KOIs in Q1-Q17 DR25 table that are not known false positives that have rotation periods
\newcommand{\nkoisnofpwithprot}{{$\sim$1{,}000}}
% fraction of KOIs that are not FPs with B+20 parameters
\newcommand{\frackoisnofpwithprotwithbtwenty}{{$\approx$92\%}}

@@

Discovering a young planet requires solving two problems: find the
planet, and measure the star's age.  Each problem admits a range of
solutions.  Here, we will focus on planets previously found using
transits, and infer new stellar ages using rotation and lithium.
Rotation will provide the most age information for the largest number
of stars.  The lithium technique will provide useful information for
stars younger than a few hundred million years.

To begin, let us imagine that the age distribution of stars within a
few hundred parsecs is uniformly distributed between 0 and 10\,Gyr
\citep[e.g.][]{Nordstrom_2004}.  
This approximiation is useful because it clarifies the rarity of young
stars:
$\approx$1\% of nearby stars have ages below 100\,Myr,
and $\approx$10\% are younger than 1\,Gyr.
Studies of exoplanets forming in disks (CITE), and of exoplanets
evolving shortly after disk dispersal (CITE CITE), are therefore
capped in their maximum achievable sample sizes by the tyranny of the
Galactic star formation rate.

@@

This work is focused on Kepler stars for which ages can be inferred
using either rotation, lithium, or both.  These ``age-dateable
stars'' are a minority of the \nkeplerstars\ Kepler targets.  Rotation
periods have been reported for roughly one in three Kepler targets
\citep[e.g.][]{McQuillan_2014,Santos_2021}.  High-resolution spectra,
and therefore measurements of the \ion{Li}{1} 6708\,\AA\ doublet, in
general have only been acquired for the Kepler objects of interest
(KOIs), which comprise a few percent of Kepler's targets.
In the following, we will describe the set of stars that we
adopt with measured rotation periods (Section~\ref{subsec:rotsel}),
the set of objects we adopt as planets
(Section~\ref{subsec:planetsel}), and the set of planets with
high-resolution spectra suitable for lithium analysis
(Section~\ref{subsec:lithiumsel}).

@@

We considered planets in the NASA Exoplanet Archive's cumulative KOI
table, which included the best knowledge available on any given planet
candidate as of 6 June 2023, while also incorporating human-based
vetting \footnote{See
\url{https://exoplanetarchive.ipac.caltech.edu/docs/PurposeOfKOITable.html}}.
These planets represent a superset of those in the fully automated
Q1-Q17 DR25 KOI Table \citep{Thompson_2018}, which could be adopted in
future work for planet occurrence rate calculations.  This version of
the cumulative KOI table included \nkoisnofp\ objects that are either
``confirmed'' or ``candidate'' planets, after excluding known false
positives.

@@

\subsection{High Resolution Spectra}
\label{subsec:lithiumsel}
A final piece of our analysis involves an age assessment based on the
equivalent width of the \ion{Li}{1} 6708\,\AA\ doublet.  For this
component of the work, we rely on spectra acquired using the High
Resolution Echelle Spectrometer (HIRES; \citealt{vogt_hires_1994}) on
the Keck I 10m telescope.  These spectra were collected between 2009
and 2024, primarily through the California Kepler Survey
\citep{2017AJ....154..107P,2017AJ....154..108J,2017AJ....154..109F}.
We supplemented the existing archive with $\approx$10~hours of new
observations for 22 stars between Fall 2022 and Spring 2024.
These stars were chosen to ensure that confirmed planets with
rotational evidence for ages below 1\,Gyr received spectra, since this
is the age range where lithium is most likely to yield useful age
constraints.

@@

Our default source for stellar temperatures and surface gravities is
the Gaia-Kepler Stellar Properties Catalog (GKSP;
\citealt{Berger_2020a_catalog}).  The GKSP parameters were reported
for stars with ``AAA'' 2MASS photometry, measured parallaxes in Gaia
DR2,  and $g$-band photometry from either the KIC or the Kepler-INT
survey.  The parameters themselves were derived using
\texttt{isoclassify} \citep{2017ApJ...844..102H} to interpolate over
the MIST isochrone grids
\citep{2016ApJ...823..102C,2016ApJS..222....8D}, given the SDSS $g$
and 2MASS $K_{\rm s}$ photometry, the Gaia DR2 parallaxes, and
$V$-band extinction from the \citet{2018MNRAS.478..651G} reddening
map.  The resulting stellar parameters are available for
\fracstarswithprotwithbtwenty\ of the \nuniqstarsantosrot\ Kepler
stars with rotation periods.  For the remaining
\fracstarswithprotwithoutbtwenty\ of stars that lack temperatures and
surface gravities from \citetalias{Berger_2020a_catalog}, we adopt the
values reported by \citet{Santos_2019} and \citet{Santos_2021}, which
for these cases primarily derive from the \citet{Mathur_2017} DR25
Kepler Stellar Properties Catalog, and are mostly derived from
photometry.  In the planet sample, \frackoisnofpwithprotwithbtwenty\
of the non-false-positive KOIs with rotation periods have parameters
from \citet{Berger_2020a_catalog}, and the remainder similarly are
drawn from DR25.

\citet{David_2021} compared the photometric
\citetalias{Berger_2020a_catalog} stellar parameters ($T_{\rm eff}$,
$R_\star$, [Fe/H]) against the spectroscopic parameters from
\citet{Fulton_2018}.  The temperature scales did show a few-percent
systematic difference, with \citetalias{Fulton_2018} quoting higher
temperatures than \citetalias{Berger_2020a_catalog} for mid K dwarfs,
and lower temperatures for early F dwarfs.  Our age analysis,
described below, assumes Gaussian uncertainties in effective
temperatures; we by default adopt the maximum of the two-sided
uncertainties reported by \citetalias{Berger_2020a_catalog} as our
symmetric temperature uncertainty.  Overall, the impact of systematic
uncertainties in the temperature scale on our age uncertainties is
generally smaller than that of the intrinsic scatter in the
calibration sequences of the open clusters.

%TODO: plot uncertainties in teff vs teff...
%TODO: plot uncertainties in logg vs logg...

%TODO: add that plot showing differences btwn the two teff scales?

@@

To evaluate the accuracy and precision of our equivalent measurement
technique, after applying an initial iteration on the
\nplhoststarwgyroagewithgrazingandhighruwe\ aforementioned stars,
we compared our lithium equivalent widths
with those reported by \citet{2018ApJ...855..115B}.
For the \nbergeroverlap\ stars in both samples, we found {\bf X, Y,
Z.}
